The Galactic Delay-Time Distribution of Type Ia Supernovae:\\ A Chemical Evolution Perspective

### 2023-06-15 src/tex/ms.tex
[showyourwork] overleaf sync

% Begin!
\begin{document}

@@

% Abstract with filler text
\begin{abstract}
    Abstract.
\end{abstract}

@@

We explore four different classes of DTD: a power-law, a broken power-law with an initial flat plateau, an exponential, and an exponential with a prompt Gaussian component, with one or two different parameterizations for each. Figure \ref{fig:dtds} presents these DTDs as a function of time since star formation.

The power-law DTD is motivated by studies of the cosmic Ia rate \citep{Maoz2017-CosmicDTD,Wiseman2021-DESRates}, and we explore slopes of $\alpha=-1.1$ and $\alpha=-1.4$ which correspond to the field galaxy and galaxy cluster DTDs of \citet{Maoz2017-CosmicDTD}, respectively. We explore the broken power-law with plateau widths of 300 Myr and 1 Gyr \textemdash the first as a proxy for both

\begin{enumerate}
    \item A simple power law with slope $\alpha$:
    \begin{equation}
        R(t) = A_\alpha (t/1\,\rm{Gyr})^\alpha
        \label{eq:powerlaw-dtd}
    \end{equation}
    where $A_\alpha = (\alpha+1)(t_{\rm max}^{\alpha+1} - t_{\rm min}^{\alpha+1})^{-1}$ is the normalization coefficient.
    We investigate the cases $\alpha=-1.1$ and $\alpha=-1.4$, following the field galaxy and galaxy cluster DTDs of \citet{Maoz2017-CosmicDTD}, respectively.

    \item A ``plateau'' model consisting of a flat slope of width $W$ followed by a power law decline with normalization $B_W$:
    \begin{equation}
        R(t) = 
        \begin{cases}
            B_W, & t < W \\
            B_W (t/W)^\alpha, & t \ge W
        \end{cases}
        \label{eq:plateau-dtd}
    \end{equation}
    where $B_W=[(W-t_{\rm min}) + \frac{t_{\rm max}^{\alpha+1} - W^{\alpha+1}}{(\alpha+1)W^\alpha}]^{-1}$ is the normalization coefficient.
    We investigate the cases $W=0.3$ Gyr and $W=1$ Gyr, taking $\alpha=-1.1$ for all plateau models. Figure \ref{fig:dtd-greggio05} illustrates the similarity between these models and two different treatments of the analytical double-degenerate DTD from \citet{Greggio2005-AnalyticalRates}.

    \item An exponentially declining DTD with timescale $\tau$ and normalization $C_\tau$:
    \begin{equation}
        R(t) = 1/\tau \exp(-t/\tau)
        \label{eq:exponential-dtd}
    \end{equation}
    We investigate the case where $\tau=1.5$ Gyr, which has been used in previous studies \citep[e.g.,][]{Schonrich2009-RadialMixing,Weinberg2017-ChemicalEquilibrium} and is similar to the analytical single-degenerate DTD from \citet{Greggio2005-AnalyticalRates}. We also investigate a longer timescale of $\tau=3$ Gyr.

    \item A DTD in which 50\% of SNe Ia belong to a ``prompt'' Gaussian component at small $t$ and the other 50\% form an exponential tail at large $t$, normalized by $D_{t_p,\sigma,\tau}$:
    \begin{equation}
        R(t) = D_{t_p,\sigma,\tau} \Big[\exp\Big(-\frac{(t-t_p)^2}{2\sigma^2}\Big) + \exp(-t/\tau)\Big]
        \label{eq:prompt-dtd}
    \end{equation}
    Following the two-population model from \citet{Mannucci2006-TwoPopulations}, we take $t_p=50$ Myr, $\sigma=15$ Myr, and $\tau=3$ Gyr, which results in roughly 50\% of SNe Ia exploding within $t<100$ Myr.

    \item A DTD based on simulations of triple-system evolution by \citet{Rajamuthukumar2022-TripleEvolution}. We approximate their DTD as a special case of the plateau model (Equation \ref{eq:plateau-dtd}) where the initial rate is quite low until an instantaneous rise to the plateau value at time $t_{\rm rise}$:
    \begin{equation}
        R(t) = 
        \begin{cases}
            \epsilon B_W, & t < t_{\rm rise} \\
            B_W, & t_{\rm rise} \leq t < W \\
            B_W (t / W) ^ \alpha, & t \geq W
        \end{cases}
        \label{eq:triple-dtd}
    \end{equation}
    We take $t_{\rm rise}=0.5$ Gyr, $W=0.5$ Gyr, $\alpha=-1.1$, and $\epsilon=0.05$ so the initial rate is 5\% of the peak rate.
\end{enumerate}

@@

\subsection{Chemical Evolution Models}
\label{sec:vice-models}

\subsubsection{Stellar Migration}
\label{sec:migration}

\begin{equation}
    \sigma_{\Delta R} = 1.35\,{\rm kpc}(R_{\rm form}/8\,{\rm kpc})^{0.61} (\tau/1\,{\rm Gyr})^{0.33}
    \label{eq:radial-migration}
\end{equation}

\begin{equation}
    {\rm CDF}(z_{\rm final}) = \frac{1}{1 + \exp(-z_{\rm final} / h_z)}.
    \label{eq:sech-cdf}
\end{equation}

\begin{equation}
    h_z = (0.25\,{\rm kpc}) 
    \exp\Big(\frac{\tau-5\,{\rm Gyr}}{7.0\,{\rm Gyr}}\Big)
    \exp\Big(\frac{R_{\rm final}-8\,{\rm kpc}}{6.0\,{\rm kpc}}\Big)
    \label{eq:scale-height}
\end{equation}

>>>>

\begin{document}

@@

% Abstract from AAS 241
\begin{abstract}
    Type Ia supernovae (SNe Ia) produce most of Fe-peak elements in the Universe and therefore are a crucial ingredient in galactic chemical evolution models. SNe Ia do not explode immediately after star formation and the delay-time distribution (DTD) has not been definitively determined by supernova surveys or theoretical models. Because the DTD also affects the relationship among age, [Fe/H], and [$\alpha$/Fe] in chemical evolution models, comparison with observations of stars in the Milky Way is an important consistency check for any proposed DTD. We implement several popular forms of the DTD in combination with multiple star formation histories for the Milky Way in one-zone and multi-zone chemical evolution models using the Versatile Integrator for Chemical Evolution (VICE). VICE includes a prescription for radial stellar migration in multi-zone models. We compare our predicted interstellar medium abundance tracks, stellar abundance distributions, and stellar age distributions to the 17th data release of the Apache Point Observatory Galactic Evolution Experiment (APOGEE). We find that the SN Ia DTD has the largest effect on the [$\alpha$/Fe] distribution: a DTD with more prompt SNe Ia produces a stellar abundance distribution that is skewed toward a lower [$\alpha$/Fe] ratio, whereas a DTD with more delayed SNe Ia produces a distribution that is skewed toward higher [$\alpha$/Fe]. While the DTD alone cannot explain the observed bimodality in the [$\alpha$/Fe] distribution, in combination with an appropriate star formation history it affects the goodness of fit between the simulated and observed high-$\alpha$ sequence.
\end{abstract}

@@

We explore five different classes of DTD: a power-law, a broken power-law with an initial flat plateau, an exponential, an exponential with a prompt Gaussian component, with one or two different parameterizations for each. Figure \ref{fig:dtds} presents these DTDs as a function of time since star formation.

\paragraph{Power-law} A simple power law with slope $\alpha$:
\begin{equation}
    R(t) = A_\alpha (t/1\,\rm{Gyr})^\alpha
    \label{eq:powerlaw-dtd}
\end{equation}
where $A_\alpha = (\alpha+1)(t_{\rm max}^{\alpha+1} - t_{\rm min}^{\alpha+1})^{-1}$ is the normalization coefficient.
We investigate the cases $\alpha=-1.1$ and $\alpha=-1.4$, which correspond to the field galaxy and galaxy cluster DTDs of \citet{Maoz2017-CosmicDTD}, respectively.

\paragraph{Plateau} A ``plateau'' model consisting of a flat slope of width $W$ followed by a power law decline with normalization $B_W$:
\begin{equation}
    R(t) = 
    \begin{cases}
        B_W, & t < W \\
        B_W (t/W)^\alpha, & t \ge W
    \end{cases}
    \label{eq:plateau-dtd}
\end{equation}
where $B_W=[(W-t_{\rm min}) + \frac{t_{\rm max}^{\alpha+1} - W^{\alpha+1}}{(\alpha+1)W^\alpha}]^{-1}$ is the normalization coefficient.
We investigate the cases $W=0.3$ Gyr and $W=1$ Gyr, taking $\alpha=-1.1$ for all plateau models. Figure \ref{fig:dtd-greggio05} illustrates the similarity between these models and two different treatments of the analytical double-degenerate DTD from \citet{Greggio2005-AnalyticalRates}.

\paragraph{Exponential} An exponentially declining DTD with timescale $\tau$ and normalization $C_\tau$:
\begin{equation}
    R(t) = \frac{1}{\tau} e^{-t/\tau}
    \label{eq:exponential-dtd}
\end{equation}
We investigate the case where $\tau=1.5$ Gyr, which has been used in previous studies \citep[e.g.,][]{Schonrich2009-RadialMixing,Weinberg2017-ChemicalEquilibrium} and is similar to the analytical single-degenerate DTD from \citet{Greggio2005-AnalyticalRates}. We also investigate a longer timescale of $\tau=3$ Gyr.

\paragraph{Prompt} A DTD in which 50\% of SNe Ia belong to a ``prompt'' Gaussian component at small $t$ and the other 50\% form an exponential tail at large $t$, normalized by $D_{t_p,\sigma,\tau}$:
\begin{equation}
    R(t) = D_{t_p,\sigma,\tau} \Big[e^{-\frac{(t-t_p)^2}{2\sigma^2}} + e^{-t/\tau}\Big]
    \label{eq:prompt-dtd}
\end{equation}
Following the two-population model from \citet{Mannucci2006-TwoPopulations}, we take $t_p=50$ Myr, $\sigma=15$ Myr, and $\tau=3$ Gyr, which results in roughly 50\% of SNe Ia exploding within $t<100$ Myr.

\paragraph{Triple} A DTD based on simulations of triple-system evolution by \citet{Rajamuthukumar2022-TripleEvolution}. We approximate their DTD as a special case of the plateau model (Equation \ref{eq:plateau-dtd}) where the initial rate is quite low until an instantaneous rise to the plateau value at time $t_{\rm rise}$:
\begin{equation}
    R(t) = 
    \begin{cases}
        \epsilon B_W, & t < t_{\rm rise} \\
        B_W, & t_{\rm rise} \leq t < W \\
        B_W (t / W) ^ \alpha, & t \geq W
    \end{cases}
    \label{eq:triple-dtd}
\end{equation}
We take $t_{\rm rise}=0.5$ Gyr, $W=0.5$ Gyr, $\alpha=-1.1$, and $\epsilon=0.05$ so the initial rate is 5\% of the peak rate.

\subsection{Star Formation Histories}
\label{sec:sfh}

\begin{figure*}
    \centering
    \includegraphics{figures/star_formation_histories.pdf}
    \caption{Caption}
    \label{fig:sfhs}
    \script{star_formation_histories.py}
\end{figure*}

\paragraph{Inside-out}
\begin{equation}
    f_{\rm IO}(t|R_{\rm gal}) = (1 - e^{-t/\tau_{\rm rise}}) e^{-t/\tau_{\rm SFH}}
    \label{eq:insideout-sfh}
\end{equation}

\paragraph{Late-burst}
\begin{equation}
    f_{\rm LB}(t|R_{\rm gal}) = f_{\rm IO}(t|R_{\rm gal}) \Big(1 + A_b e^{-(t-t_b)^2/2\sigma_b^2} \Big)
\end{equation}

\paragraph{Early-burst} The infall rate at a given radius declines exponentially with time:
\begin{equation}
    f_{\rm EB}(t|R_{\rm gal}) = e^{-t/\tau_{\rm IFR}}
    \label{eq:earlyburst-ifr}
\end{equation}
We take $\tau_{\rm IFR} = \tau_{\rm SFH}$ at all radii.
The time-dependence of the SFE timescale $\tau_*$ comes from \citet{Conroy2022-ThickDisk}:
\begin{equation}
    \frac{\tau_*}{1\,\rm{Gyr}} =
    \begin{cases}
        50 & t < 2.5\,\rm{Gyr} \\
        50/[1+3(t-2.5)]^2 & 2.5\leq t \leq 3.7\,\rm{Gyr} \\
        2.36 & t > 3.7\,\rm{Gyr}
    \end{cases}
    \label{eq:earlyburst-taustar}
\end{equation}

\paragraph{Two-infall}

\subsection{Stellar Migration}
\label{sec:migration}

\begin{equation}
    \sigma_{\Delta R} = 1.35\,{\rm kpc} \Big(\frac{R_{\rm form}}{8\,{\rm kpc}}\Big)^{0.61} \Big(\frac{\tau}{1\,{\rm Gyr}}\Big)^{0.33}
    \label{eq:radial-migration}
\end{equation}

\begin{equation}
    {\rm CDF}(z_{\rm final}) = \frac{1}{1 + e^{-z_{\rm final} / h_z}}.
    \label{eq:sech-cdf}
\end{equation}

\begin{equation}
    h_z = (0.25\,{\rm kpc}) 
    e^{\frac{\tau-5\,{\rm Gyr}}{7.0\,{\rm Gyr}}}
    e^{\frac{R_{\rm final}-8\,{\rm kpc}}{6.0\,{\rm kpc}}}
    \label{eq:scale-height}
\end{equation}

@@
